---
title: "Case Study: Weekend Drinking (An Intro to ZIP Models)"
format: docx
---

## Load Libraries

```{r}
library(tidyverse)
library(pscl) # functions for zero inflated Poisson models
```

## Read in Data

An informal survey of students in an introductory statistics course included the question, "How many alcoholic drinks did you consume last weekend?". This survey was conducted on a dry campus where no alcohol is officially allowed, even among students of drinking age.

```{r}
zip.data = read.csv("data/weekendDrinks.csv") |> rename(gender = sex)
```

Each line of `weekendDrinks.csv` contains data provided by a student in an introductory statistics course. In this analysis, the response of interest is the respondent's report of the number of alcoholic `drinks` they consumed the previous weekend, whether the student lives `off.campus`, and `gender` (labeled `sex` in the data file). We will also consider if a student is a `firstYear` student.

## Exploratory Data Analysis

**Task: Modify the code for "gender.table" to make similar tables summarizing `off.campus` and `firstYear`.**

```{r}
gender.table <- zip.data |>
              group_by(gender) |>
              tally() |>
              mutate(prop = round(n/sum(n),3) )
gender.table



off.table <- zip.data |>
              group_by(off.campus) |>
              tally() |>
              mutate(prop = round(n/sum(n),3) )
off.table


fy.table <- zip.data |> 
            group_by(firstYear) |>
            tally() |>  
            mutate(prop = round(n/sum(n),3))
fy.table
```

This is a histogram of our response variable of interest (number of drinks)

```{r}
ggplot(zip.data, aes(x=drinks)) + 
  geom_histogram(fill="hotpink3", color="black", bins=15) + 
  xlab('Number of Drinks "Last" Weekend') 

```

This is a histogram of data from a Poisson distribution with the same mean as the response variable from our dataset (and has the same number of observations as our dataset).

```{r}
set.seed(925203)
pdat = data.frame(fish=rpois(77, mean(zip.data$drinks)))  # notice the lame joke here?
ggplot(pdat, aes(x=fish))+
  geom_histogram(fill="thistle4", color="black", bins=15)
```

**Task: Compare and contrast these two histograms. How are they similar? How are they different?**

**Task: Come up with at least two reasons why a student might have had 0 drinks "last" weekend.**

**Task: Modify the code for gender to create similar code for off.campus and firstYear.**

```{r}
zip.data |>
  group_by(gender) |>
  summarise(n=n(),
            mean=mean(drinks),
            var = var(drinks)
  )

zip.data |>
  group_by(off.campus) |>
  summarise(n=n(),
            mean=mean(drinks),
            var = var(drinks)
  )

zip.data |>
  group_by(firstYear,gender) |>
  summarise(n=n(),
            mean=mean(drinks),
            var = var(drinks)
  )

```

**Task: What do the means and variances in the above summaries suggest about the Poisson regression mean = variance assumption?**

```{r}
ggplot(zip.data, aes(x=gender, y=drinks, fill=as.factor(off.campus))) + geom_boxplot() + ylab('Number of drinks "last" weekend')

ggplot(zip.data, aes(x=gender, y=drinks, fill=firstYear)) + geom_boxplot() + ylab('Number of drinks "last" weekend')

ggplot(zip.data, aes(x=firstYear, y=drinks, fill=as.factor(off.campus))) + geom_boxplot() + ylab('Number of drinks "last" weekend')
```

# Stop here in prep for Wednesday's task

## Poisson Regression Model

```{r}
PRmod = glm(drinks ~ gender + off.campus, data=zip.data, family="poisson")
summary(PRmod)
exp(coef(PRmod))
```

**Provide interpretations of the estimated coefficients in the Poisson Regression model.**

## Zero-Inflated Poisson Regression Model

```{r}
ZIPmod = zeroinfl(drinks ~ gender + off.campus | firstYear,    # pscl package
         data = zip.data,
         dist = "poisson")

summary(ZIPmod)
exp(coef(ZIPmod))
```

**Provide interpretations of the estimated coefficients in the Poisson Regression portion of the ZIP model.**

**Predict the probability that a first-year student is a non-drinker. Predict the probability that a non-first year student is a non-drinker.**

$\eta$

## Comparing ZIP and Ordinary Poisson Regression

```{r}
vuong(PRmod, ZIPmod)
```

Vuong test indicates that the ZIP model is a better fit for these data than a regular Poisson regression model. (as indicated by the small p-values)


## Tweak Zip model to use gender and off.campus as predictors of non-drinker status

```{r}
ZIPmod2 = zeroinfl(drinks ~ gender + off.campus | gender + off.campus,    # pscl package
         data = zip.data,
         dist = "poisson")

summary(ZIPmod2)
```
## Tweak Zip model to not use any predictors of non-drinker status

"Intercept only model" for the zero-inflation

```{r}
ZIPmod3 = zeroinfl(drinks ~ gender + off.campus | 1 ,    # pscl package
         data = zip.data,
         dist = "poisson")

summary(ZIPmod3)
```













