---
title: "Binary Logistic Regression"
format: docx
---

## Load Libraries

```{r}
library(tidyverse)
```

## Read in Data

```{r}
movies = read_csv("data/movies90.csv")
```

IMDb, run by Amazon, claims to be the largest movie database in the world. In this dataset, there are 2783 that debuted between 1990 and 2005. Relevant variables in the dataset include \* `Drama` = an indicator variable, 1 if the film is a drama and 0 otherwise \* `rating` = the movie's rating on IMDb (as of several years ago) \* `length` = the movie's length (in minutes) \* `budget` = the movie's budget

In this analysis, we will look for predictors of a movie being a `drama`.

# Exploratory Data Analysis

**Response Variable**

```{r}
ggplot(movies, aes(x=Drama)) +
  geom_bar(fill="wheat1") + 
  xlab("Drama Indicator (1=yes, 0=no)") +
  theme_dark()
```

**Explore Rating as a potential Predictor**

```{r}
# Explore Rating as a potential Predictor
sumStats_rating <- movies |>
  group_by(rating = ntile(rating, 278)  ) |> # group movies with similar ratings together
  summarise(n = n(),
            p.est = mean(Drama), # proportion of dramas
            emp.odds = p.est/(1-p.est), # estimated odds
            emp.logit = log(emp.odds)) # estimated logit (aka log-odds)

ggplot(data = sumStats_rating, aes(x=rating, y=emp.logit)) + 
  geom_point() + 
  geom_smooth(method="loess") + 
  geom_hline(yintercept=0, color="red")

```

**Explore Length as a potential Predictor**

```{r}
# Explore Length as a potential Predictor
sumStats_length <- movies |>
  group_by(length = ntile(length, 500)) |>
  summarise(n=n(),
            p.est = mean(Drama),
            emp.odds = p.est/(1-p.est),
            emp.logit = log(emp.odds))

ggplot(data = sumStats_length, aes(x=length, y=emp.logit)) + 
  geom_point() + 
  geom_smooth(method="loess") + 
  geom_hline(yintercept=0, color="red")

```

**Explore Budget as a potential Predictor**

```{r}
# Explore Budget as a potential Predictor
sumStats_budget <- movies |>
  group_by(budget) |>
  summarise(n=n(),
            p.est = mean(Drama),
            emp.odds = p.est/(1-p.est),
            emp.logit = log(emp.odds))

ggplot(data = sumStats_budget, aes(x=budget, y=emp.logit)) + 
  geom_point() + 
  geom_smooth(method="loess") + 
  geom_hline(yintercept=0, color="red")

```

# Fit Logistic Regression Models

```{r}
mod1 = glm(Drama ~ rating, family="binomial", data=movies)
summary(mod1)
exp(coefficients(mod1))
confint(mod1)  # log-odds scale
confint(mod1) |> exp() # odds scale
```

**Model 1 Questions**

1.  Is rating a significant predictor?
2.  Provide an interpretation of the estimated coefficient on rating.
3.  Construct and interpret a confidence interval for the population coefficient associated with IMDb rating.

```{r}
exp(c(0.3369, 0.4529))
```

4.  Is there evidence of model lack of fit?

```{r}
summary(mod1)

pchisq(3653, df = 2781, lower.tail = FALSE)

```
Yes, there is a lack of fit issue here.


### Model 2: Quadratic effects

```{r}
mod2 = glm(Drama ~ rating + I(rating^2) + length + I(length^2) + budget, family="binomial", data=movies)
summary(mod2)
```

**Model 2 Questions**

1.  Is there evidence that the model is useful?
2.  Is there evidence that the terms involving length significantly improve the model? (Address with a single hypothesis test.)

```{r}
mod2b <- glm( Drama ~ rating + I(rating^2) + budget, family = "binomial", data = movies)
summary(mod2b)
```

3.  Is there evidence that the quadratic terms significantly improve the model? (Address with a single hypothesis test.)

```{r}

mod2c <- glm(Drama ~ rating +  length + budget, family="binomial", data=movies)
summary(mod2c)

anova(mod2c, mod2, test = "Chisq")
```

4.  Is budget useful in the model?

```{r}
summary(mod2)
```


5.  Provide an interpretation of the estimated coefficient on budget.
```{r}
coef(mod2) |> exp()
```
Shift to one million dollar increase

```{r}
1000000 * coef(mod2)

exp(-0.0208595)
```

```{r}
movies <- movies |>
            mutate(budget_millions = budget/1000000)

mod2d <- glm(Drama ~ rating + I(rating^2) + length + I(length^2) + budget_millions, 
             family="binomial", data=movies)

summary(mod2d)

```




6.  Is there evidence of model lack of fit?

```{r}
summary(mod2)
pchisq(3214.1, 2777, lower.tail = FALSE)
```





