---
title: "Introduction to Poisson Regression"
format: docx
---

## Load Libraries

```{r}
library(tidyverse)

source("code_initial_version/plotPoissonDevRes.R")
```

## Load Philippines Household data

```{r}
fHH1 = read_csv("data/fHH1.csv")
```

## Exploratory Data Analysis

Explore the response variable
```{r}
ggplot(fHH1, aes(x = total)) + 
  geom_histogram(binwidth = .25) + 
  xlab("Number in the house (excluding\nhead of household)") +
  ylab("Count of households")

mean(fHH1$total)
var(fHH1$total)
```

Explore Summary Stats by Roof Material
```{r}

table(fHH1$roof) |> prop.table() |> signif(3)

fHH1  |>
  group_by(roof)  |>
  summarise(n = n(),
            mean = mean(total), 
            var = var(total)
            )

```

Explore Summary Stats by location
```{r}
fHH1  |>
  group_by(location)  |>
  summarise(n = n(),
            mean = mean(total),  
            var = var(total)
            )
```

Explore by age
```{r}

sumStats <- fHH1 |>
  group_by(age) |> 
  summarise(n = n(),
            mntotal = mean(total),
            vartotal = var(total), 
            logmntotal = log(mntotal)
            )

ggplot(sumStats, aes(x = age, y = logmntotal)) +
  geom_point() +
  geom_smooth(method = "loess", size = 1.5) +
  xlab("Age of head of the household") +
  ylab("Log of the empirical mean number in the house") 

```

Check Mean = Variance assumption
```{r}
# check mean = variance
ggplot(sumStats, aes(x = mntotal, y = vartotal)) + 
  geom_point() +
  geom_smooth(method = "loess", size = 1.5) + 
  geom_abline(slope = 1, intercept = 0,
              col = "tomato", size = 1, linetype = 3) +
  xlab("Empirical mean number in the house (by age)") +
  ylab("Empirical variance in number in the house (by age)") 

```


Explore by age and location
```{r}
sumStats2 <- fHH1 |> 
  group_by(age, location) |>
  summarise(n = n(),
            mntotal = mean(total),
            vartotal = var(total),
            logmntotal = log(mntotal)
            )

ggplot(sumStats2, aes(x = age, y = logmntotal, color = location)) +
  geom_point()+
  geom_smooth(method = "loess", se=FALSE)+
  xlab("Age of head of the household") +
  ylab("Log empirical mean household size")

```


Explore by Age and Roof Material
```{r}
sumStats3 <- fHH1 |>
  group_by(age, roof) |> 
  summarise(n = n(),
            mntotal = mean(total),
            vartotal = var(total),
            logmntotal = log(mntotal)
            )

ggplot(sumStats3, aes(x=age, y=logmntotal, color=roof)) +
  geom_point() +
  geom_smooth(method = "loess", se=FALSE) +
  xlab("Age of head of the household") +
  ylab("Log empirical mean household size") +
  theme(legend.position = "bottom")

```



## Poisson Regression Models


1.	Propose and fit a Poisson regression model that uses age to predict mean total number in household.
```{r}
mod1 = glm(, family= , data=fHH1)
summary(mod1)
confint(mod1)
```


a.	Is there evidence that age is a useful predictor in the model?
b.  Provide an interpretation of the estimated coefficient on age, in context.
c.	Construct and interpret a 95% confidence interval for population coefficient on age.
d.  Explore the poisson regression deviance residuals plot.

```{r}
plotPoissonDevRes(mod1)
```


# Prep for Wednesday Task: Finish 2 a - d
2.	Propose and fit a Poisson regression model that uses a quadratic function of age to predict mean total number in household.

```{r}
mod2 = glm(data=fHH1)


```


a.	Is there evidence that the quadratic term is useful in the model?
b.  Compute a 95% confidence interval for the relative risk associated with the quadratic term.
c.  Does the Poisson deviance residuals plot suggest this model has better fit?
d.  Compare AIC for Model 1 and Model 2? Which model does AIC suggest is better?

# Stop Here!
e.	Is there evidence that the overall model is useful?




3.	Propose and fit a Poisson regression model that uses a quadratic function of age, as well as location and roof material to predict the mean total number in the household.

a.	Is there evidence that the overall model is useful?
b.  Is there evidence that roof material is useful in the model?

```{r}
mod3 = glm(total ~ age + I(age^2) +  , family= , data=fHH1)
summary(mod3)

```




4.	Propose and fit a Poisson regression model that uses location and a quadratic function of age to predict the mean total number in the household.

```{r}
mod4 = glm(total ~ age + I(age^2) + location, family="poisson", data=fHH1)
summary(mod4)

```

a.	Is there evidence that the location terms significantly improve the model? Address this question with a single hypothesis test.

```{r}
anova(mod2,mod4, test="Chisq")
```


```{r}
fHH1 = mutate(fHH1, locationfac = as.factor(location))
mod4 = glm(total ~ age + I(age^2) + locationfac, family="poisson", data=fHH1)


library(multcomp)
summary(glht(mod4, mcp(locationfac="Tukey")))

```



