---
title: "Introduction to Multilevel Models"
format: docx
---


## Load Libraries
```{r}
library(tidyverse)
library(lme4) # for fitting multilevel models
```

## Load Data
```{r}
musicdata = read_csv("data/musicdata.csv")
```

## Understand the structure of the music dataset
* Identify response variable
* Identify Level One observational units
* Identify Level Two observational units
* Identify Level One covariates
* Identify Level Two covariates


## Exploratory Data Analysis
```{r}
ggplot(musicdata,aes(x=previous,y=na)) + 
  geom_point() + 
  geom_smooth(method="loess") + 
  xlab("# of Previous Performances") + 
  ylab("na (Negative Affect)")
```

```{r}
ggplot(musicdata,aes(x=previous,y=na)) + 
  geom_point() + 
  geom_smooth(method="lm",se=FALSE) +  # or loess?? 
  facet_wrap( ~id) +
  xlab("# of Previous Performances") + 
  ylab("na (Negative Affect)")
```

Describe the relationship between number of previous performances and negative affect for Musicians 1, 27, 38, and 40.



# Multilevel Models

## Random Intercept model


### Write out the multilevel model

-   Level One: $Y_{ij} = a_{i}+\epsilon_{ij}$ where $\epsilon_{ij}\sim N(0,\sigma^2)$

-   Level Two: $a_{i} = \alpha_{0}+u_{i}$ where $u_{i}\sim N(0,\sigma_{u}^{2})$

-   $Y_{ij}$ is the Performance anxiety scores of subject $i$ for performance $j$.

-   $a_{i}$ is the true mean response of all observations for subject $i$.

-   $\alpha_{0}$ is the grand mean -- the true mean of all observations across the entire population.

-   $\sigma^2$ is the within-person variability

-   $\sigma_{u}^{2}$ is the between-person variability

Can combine all of these into a composite model: 




The name **random intercepts model** then arises from the Level Two equation for $a_{i}$: each subject's intercept is assumed to be a random value from a normal distribution centered at $\alpha_{0}$ with variance $\sigma_{u}^{2}$.



### Fit the model

```{r, comment=NA}
#Model A (Unconditional means model)
model.a <- lmer(na ~ 1 + (1 | id), REML = T, data = musicdata)
summary(model.a)
```

### Interpret multilevel model




## Multilevel model with convariates

### Write out the multilevel model

-   $\hat{\alpha}_{0}=$

-   $\hat{\sigma}^2=$

-   $\hat{\sigma}_{u}^{2}=$


### Fit the model

```{r}
mlmod1 = lmer(na ~ previous + (previous|id), REML=T, data=musicdata)
summary(mlmod1)
```

## Interpret multilevel model


