---
title: "Multilevel Models, Part 2"
author: "Stat 313 (Chapman)"
date: "10/20/2021"
output: word_document
---

## Load Libraries
```{r}
library(tidyverse)
library(lme4) # for fitting multilevel models
```


## Load Data
```{r}
musicdata = read_csv("data/musicdata.csv")
```

## Our Current Model
```{r}
mlmod1 = lmer(na ~ previous + (previous|id), REML=T, data=musicdata)
summary(mlmod1)
```

## Why might musicians have different intercepts/slopes? 

If we want to try to explain why different musicians have different intercepts and slopes, we might try to use a Level Two covariate (predictors whose values are recorded on musicians, rather than performances.) 

### Exploratory Data Analysis
Note that values of a Level Two covariate would be repeated for each performance by a musician. This means there is no perfect way to approach the exploratory data analysis. Our options: Use every data point for a musician or summarize the data for a musician.

__Numerical Predictors (at Level Two)__

_Using Every Data Point for  Musician_

Explore the possible numerical predictors.
```{r}
ggplot(data=musicdata,aes(x=mpqpem,y=na)) + 
  geom_point() + 
  geom_smooth(method="lm",color="black") + 
  ylab("Negative Affect") + 
  xlab("PEM") 

ggplot(data=musicdata,aes(x=mpqnem,y=na)) + 
  geom_point() + 
  geom_smooth(method="lm",color="black") + 
  ylab("Negative Affect") + 
  xlab("NEM") 

ggplot(data=musicdata,aes(x=mpqab,y=na)) + 
  geom_point() + 
  geom_smooth(method="lm",color="black") + 
  ylab("Negative Affect") + 
  xlab("Absorption") 
```


_Summarizing the Data for a Musician_

Need to reduce the data down to one observation per musician. We will use mean negative affect.
```{r}
music.lev2 <-  musicdata %>%
  group_by(id) %>%
  filter(row_number() == 1) %>%
  select(id, gender:mpqnem)

meanbysubj <- musicdata %>% group_by(id) %>%
  summarise(meanbysubj = mean(na, na.rm = TRUE))
music.lev2 <- music.lev2 %>%
  left_join(meanbysubj, by = "id")

```

Now explore the possible numerical predictors.
```{r}
ggplot(data=music.lev2,aes(x=mpqpem,y=meanbysubj)) + 
  geom_point() + 
  geom_smooth(method="lm",color="black") + 
  ylab("Mean Negative Affect") + 
  xlab("PEM") 

ggplot(data=music.lev2,aes(x=mpqnem,y=meanbysubj)) + 
  geom_point() + 
  geom_smooth(method="lm",color="black") + 
  ylab("Mean Negative Affect") + 
  xlab("NEM") 

ggplot(data=music.lev2,aes(x=mpqab,y=meanbysubj)) + 
  geom_point() + 
  geom_smooth(method="lm",color="black") +
  ylab("Mean Negative Affect") + 
  xlab("Absorption") 
```


__Categorical Predictors (at Level Two)__

_Using Every Data Point for  Musician_

Explore the possible categorical predictors.

```{r}
ggplot(data=musicdata,aes(x=gender,y=na)) + 
  geom_boxplot(fill="mediumseagreen") + 
  ylab("Negative Affect") + 
  xlab("Gender") 

ggplot(data=musicdata,aes(x=instrument,y=na)) + 
  geom_boxplot(fill="mediumseagreen") + 
  ylab("Negative Affect") + 
  xlab("Instrument") 

```


_Summarizing the Data for a Musician_
Using mean negative affect for each musician, explore the possible categorical predictors.
```{r}
ggplot(data=music.lev2,aes(x=gender,y=meanbysubj)) + 
  geom_boxplot(fill="mediumseagreen") + 
  ylab("Negative Affect") + 
  xlab("Gender") 

ggplot(data=music.lev2,aes(x=instrument,y=meanbysubj)) + 
  geom_boxplot(fill="mediumseagreen") + 
  ylab("Negative Affect") + 
  xlab("Instrument") 
```

## Fitting Models with Level Two Covariates

__Using a Numerical Level Two Covariate__

First, use a musician's negative emotionality score to "explain" differences in intercepts (baseline negative affect).
```{r}
mlmod2a = lmer(na ~ mpqnem + previous +  (previous | id), data=musicdata, REML=T)

summary(mlmod2a)
```

Now, use a musician's negative emotionality score to "explain" differences in both intercepts and slopes.
```{r}
mlmod2b = lmer(na ~ mpqnem + previous + mpqnem:previous + (previous | id), data=musicdata, REML=T)

summary(mlmod2b)
```


Center the numerical predictor to improve (make easier!) the interpretations
```{r}
musicdata = mutate(musicdata, cmpqnem = mpqnem-mean(mpqnem))
```

```{r}
mlmod2a.2 = lmer(na ~ cmpqnem + previous +  (previous | id), data=musicdata, REML=T)

summary(mlmod2a.2)

mlmod2b.2 = lmer(na ~ cmpqnem + previous + cmpqnem:previous + (previous | id), data=musicdata, REML=T)

summary(mlmod2b.2)
```


__Using a Categorical Level Two Covariate__
```{r}
# add indicator variables for two of the three instrument catgories
musicdata = musicdata %>% 
  mutate(vocal = ifelse(instrument=="voice", 1, 0),
         orch = ifelse(instrument=="orchestral instrument", 1, 0))
```

First, use a musician's instrument to "explain" differences in intercepts (baseline negative affect).
```{r}
mlmod3a = lmer(na ~ previous + vocal + orch  + (previous | id), data=musicdata, REML=T)

summary(mlmod3a)
```


Now, use a musician's instrument to "explain" differences in intercepts and slopes.
```{r}
mlmod3b = lmer(na ~ previous + vocal + orch + previous:vocal + previous:orch + (previous | id), data=musicdata, REML=T)

summary(mlmod3b)
```
