---
title: 'Multilevel Models: Comparing Models with AIC, BIC, and Likelihood Ratio Tests'
format: docx
---

## Load Libraries
```{r}
library(tidyverse)
library(lme4)
```

## Read Data
```{r}
musicdata = read_csv("data/musicdata.csv")


musicdata = mutate(musicdata, cmpqnem = mpqnem-mean(mpqnem)) # center NEM covariate
# add indicator variables for two of the three instrument categories
musicdata = musicdata %>% 
  mutate(vocal = ifelse(instrument=="voice", 1, 0),
         orch = ifelse(instrument=="orchestral instrument", 1, 0))
```


## Summarize Models from class

Model 1: Previous Performances at Level One; No predictors at Level Two

```{r}
mlmod1 = lmer(na ~ previous + (previous|id), REML=F, data=musicdata)
summary(mlmod1)
```


Model 2a: Previous Performances at Level One; (centered) NEM to explain different intercepts at Level Two
```{r}
mlmod2a.2 = lmer(na ~ cmpqnem + previous +  (previous | id), data=musicdata, REML=F)

summary(mlmod2a.2)
```

Model 2b: Previous Performances at Level One; (centered) NEM to explain different intercepts and different slopes at Level Two
```{r}
mlmod2b.2 = lmer(na ~ cmpqnem + previous + cmpqnem:previous + (previous | id), data=musicdata, REML=F)

summary(mlmod2b.2)
```

Model 3a: # Previous Performances at Level One; Instrument to explain different intercepts at Level Two
```{r}
mlmod3a = lmer(na ~ previous + vocal + orch  + (previous | id), data=musicdata, REML=F)

summary(mlmod3a)
```



```{r}
anova(mlmod1, mlmod2b.2, test="chisq")
```


